Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-04 talk/dls2014/paper/paper.tex
move around sections and remove Jython from evaluation

\subsection{Memory Requirements\label{sub:memoryreq}}

@@

Third, since we do copy on write, if all objects were to be written to
at the same time, all pages would need to be privatised for all
objects in the old object space. In that case we would need the total
amount of memory required by old objects multiplied by $N+1$
(incl. the sharing segment). This is certainly the worst case and
so far not observed in practice. During major collections we re-share
privatised pages that are currently not modified by any transaction.

@@

We evaluate our system in a Python interpreter called
PyPy\footnote{www.pypy.org} (version 2.3). PyPy is an implementation of an
interpreter for the Python language. It has a special focus on speed,
as it provides a just-in-time (JIT) compiler to speed up applications
running on top of it. For comparison, we compare between normal PyPy
using a GIL and a PyPy with STM, as well as between other Python interpreters:
\begin{description}
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\item[Jython] (version 2.7b1) is an implementation of Python on top of
  the Java Virtual Machine (JVM). Instead of a GIL, this interpreter
  uses fine-grained locking for synchronisation. This enables true
  parallelism when executing code on multiple threads. In addition, its
  integration with the JVM provides it with a JIT compiler for faster
  execution.
\end{description}

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{sec:performance-bench}.
\cfbolz{the details about why the JIT confuses everything should go into
Section~\ref{sec:performance-bench}. here just write "because the integration
with the JIT is incomplete"}
We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT~\cite{cfbolz09} bases its decision what to
compile on runtime profiling, running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

Overall, we believe that disabling it on all benchmarks except the
performance benchmarks in Section~\ref{sec:performance-bench} is better
because we can minimise non-determinism. We also do not want to depend
on the capabilities of the JIT in these experiments.

@@

\cfbolz{can you explain how you measure the errors?}

@@

@@

In this section we want to look at the general performance of our
system. As explained above, the JIT does not simply speed up the
performance by a certain factor. In order to better understand the
behaviour of our system, we therefore split this section into two
parts. We first look at how it behaves without the JIT, and then with
the JIT. We look at six benchmarks:

\begin{itemize}
\item \emph{btree} and \emph{skiplist}, which are both inserting,
  removing, and finding elements in a data structure
\item \emph{threadworms}, which simulates worms walking on a grid in
  parallel and checking for collisions with each other
\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
  all perform simple, independent computations in parallel
\end{itemize}

We use coarse-grained locking for the first three benchmarks and no
locking for the last three. This is important because Jython, which
uses fine-grained locking instead of a GIL, is only expected to scale
with the number of threads for the latter group. It is not able to
scale when using coarse-grained locking. STM, however, uses atomic
blocks instead of a single lock to synchronise accesses to the
shared data structures. Since atomic blocks map to transactions,
our STM system may still be able to get a speedup on more than one
thread by optimistically running transactions in parallel.

% To isolate factors we look at performance w/o JIT and perf w JIT.
% w/o JIT:
%  - it scales
%  - slower than reference CPython
%  - gil scales negatively
% w JIT:
%  - it sometimes still scales (more work needed)
%  - changed working set size because generally ~50x faster
%  - competing w our own JIT on non-stm a challenge
%  - gil scales negatively

\subsection{Non-JIT Benchmarks}
First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in Figure~\ref{fig:performance-nojit}.

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see~\cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
raytrace, richards}), and behave similar to the GIL interpreters in the
other cases. The reason again being the coarse grained locking.

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best in the ones where
Jython scales as well and a little less in the others. The reason
for that is that in the former group there are no real, logical
conflicts -- all threads do independent calculations. In the latter
case, the threads work on a common data structure and therefore
create much more conflicts, which limits the scalability.

@@

\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
however, it does not consistently beat CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. So we will now look at how well our system works when we
enable the JIT.

@@

\subsection{JIT Benchmarks}
The speedups from enabling the JIT in
these benchmarks range from $10-50\times$. This is why we had to do
without Jython and CPython here, since they would be much further up
in the plots. Also, in order to get more stable results, we increased
the input size of all benchmarks to get reasonable execution times.

@@

We see that generally, the group of non-locked benchmarks scales
best. The other three scale barely or not at all with the number of
threads. The reason for this is likely again the conflicts in the
latter group. Since the JIT accelerates all code but not the STM
overhead, we do more work per transaction. And this increases the
likelihood of conflicts between them and therefore limits scalability
even more than in the no-JIT benchmarks.

Overall PyPy needs the JIT in order for its performance to be
competitive. It would be interesting to see how using our STM system
in CPython would turn out, but it is a lot of work. On its own, our
system scales well, so we hope to also see that with the JIT in the
future.

@@

@@

\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
  in Parallelising Dynamic Languages. ICOOOLPS'14.

@@

>>>>

\subsection{Memory Requirements\label{sub:memoryreq}}
\remi{maybe its own ``limitations'' section? also about
  static segmentation and limited scaling }

@@

Third, since we do copy on write, if all objects were to be written to
at the same time, all pages would need to be privatised for all
objects in the old object space. In that case we would need the total
amount of memory required by old objects multiplied by $N+1$
(incl. the sharing segment). This is certainly the worst case and
so far not observed in practice. During major collections we re-share
privatised pages that are currently not modified by any transaction,
thereby reclaiming some of this memory.

@@

We evaluate our system in a Python interpreter called
PyPy\footnote{www.pypy.org} (version 2.3). PyPy is an implementation of an
interpreter for the Python language. It has a special focus on speed,
as it provides a just-in-time (JIT) compiler to speed up applications
running on top of it. For comparison, we compare between normal PyPy
using a GIL and a PyPy with STM, as well as to CPython.
\begin{description}
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\end{description}

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{subsec:jit-benchs}. The STM-JIT
integration is currently still incomplete and not tested much. The
JIT-less interpreter provides a much more consistent environment for
the STM system, so we remove some unknown variables by disabling it.

@@

For each point in the plots, we took 5 measurements and report the
average and the standard deviation as error bars. For the JIT
benchmarks, we first let warm up the JIT by doing a few
additional iterations to get a single measurement.

@@

For the following sections we use a set of six small benchmarks
available on \cite{pypybenchs}:

\begin{itemize}
\item \emph{btree} and \emph{skiplist}, which are both inserting,
  removing, and finding elements in a data structure from multiple
  threads
\item \emph{threadworms}, which simulates worms walking on a grid in
  parallel and checking for collisions with each other
\item \emph{mandelbrot}, \emph{raytrace}, and \emph{richards}, which
  all perform some independent computations in parallel (embarrassingly
  parallel)
\end{itemize}

We use atomic blocks in order to easily synchronise access to shared
data structures in the first three benchmarks. For the embarrassingly
parallel benchmarks, we do not need any explicit synchronisation at
all. These atomic blocks are simulated with a single global lock
when running on top of GIL-supported interpreters. With our STM
system, they map to transactions that will execute optimistically
in parallel.

@@

\subsection{Non-JIT Benchmarks}
First we run our benchmarks on three different interpreters: CPython
(GIL), and PyPy with STM and with the GIL (both without the JIT). The
results are shown in Figure~\ref{fig:performance-nojit}.

As expected, GIL-supported interpreters do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see~\cite{beazley10} for a detailed
analysis).

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones and a little less in the others. The reason
for that is that in the former group there are no real, logical
conflicts -- all threads do independent calculations. STM simply
replaces the GIL in those programs. In the latter group,
the threads work on a common data structure and therefore
create much more conflicts, which limits the scalability. Here
we make active use of the STM-supported atomic blocks that
provide a scalable synchronisation mechanism.

@@

\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
however, it rarely beats CPython's single-thread performance.  For
programs that need concurrency in CPython and that use threads to
achieve this, it also makes sense to look at the overhead induced by
the GIL on multiple threads. From this perspective, the STM
implementation beats CPython's performance in all but two benchmarks.

Since PyPy comes with a JIT to make its overhead compared to CPython
go away, we will now look at how well STM works together with it.

@@

\subsection{JIT Benchmarks\label{subsec:jit-benchs}}

We would like to regard enabling the JIT as a simple performance
enhancement, but that is not what happens in reality. First, since the
JIT~\cite{cfbolz09} bases its decision of what to compile on runtime
profiling, running in multiple threads, it may compile different
things in each run because of the non-deterministic thread-scheduling
of the operating system (OS). Second, it is able to remove some
allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.
For these reasons, the following results have to be taken with a grain
of salt.

The speedups from enabling the JIT in these benchmarks range from
$10-50\times$. This is why we had to do without CPython here, since it
would be much further up in the plots. Also, in order to make jitting
code worthwhile, we increased the input size of all benchmarks to get
reasonable execution times.

We also ran these benchmarks on Jython\footnote{version
2.7b1~\cite{webjython}}, an implementation of Python on top of the
Java Virtual Machine (JVM).  Instead of a GIL, this interpreter uses
fine-grained locking for synchronisation. Even though it can use the
JVM's JIT compiler, its performance in these benchmarks is behind
PyPy's by a factor of \remi{$5-10\times$}. Because of that it does not
make sense to include it in this evaluation. In the future, we would
like to do an extensive comparison between the STM approach and
fine-grained locking. It is out of the scope of this paper to do
this thoroughly.

@@

We see that generally, the group of embarrassingly parallel benchmarks scales
best. The other three scale barely or not at all with the number of
threads. The reason for this is likely again the conflicts in the
latter group. Since the JIT accelerates all code but not the STM
overhead, we do more work per transaction. This increases the
likelihood of conflicts between them and therefore limits scalability
even more than in the no-JIT benchmarks.

Overall PyPy needs the JIT in order for its performance to be
competitive with CPython's. It would be interesting to see how using
our STM system in CPython would turn out, but it is a lot of work. On
its own, our system scales well, so we hope to also see that with the
JIT in the future.

@@

%%% TODO: fix all citation to follow common format (always page numbers; no 'et al', etc)

@@

\bibitem{wayforward14} Remigius Meier, Armin Rigo. 2014. A Way Forward
  in Parallelising Dynamic Languages. To appear in ICOOOLPS'14.

@@

\bibitem{pypybenchs} PyPy benchmarks repository. 2014. Revision
  a26f2fb58413. \url{bitbucket.org/pypy/benchmarks}
